---
title: "Data preparation, exploratory data analysis and model evaluation with funModeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data preparation, exploratory data analysis and model evaluation with funModeling}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, message=FALSE}
library(funModeling)
```


## Exploratory Data Analysis

### Using `freq` function to profile categorical variables

```{r, fig.width=3, fig.height=2}
vars_to_process=c("thal", "chest_pain")
freq(heart_disease, vars_to_process)
```

It accepts a single vector. Also, if parameter `path_out` is added it exports every plot in high quality. If the folder doesn't exist, it will be created.

```{r, eval=FALSE}
freq(heart_disease$thal, path_out = "the_best_folder")
```

If str_input is not specified, then it runs for all factor and character variables

```{r, eval=FALSE}

freq(heart_disease)
```

## Selecting best variables (dataViz)

There are 2 functions that can help us in the visualization of the input variables in regarding the target variable:

`cross_plot` and `plotar`.

Both, just the same as `freq` variable:

* Can handle from 1 to N input variables, using the `str_input` parameter.
* If `str_input` is not specified, then they will run for all variables present in the data.
* Can export all the generated plots into a folder, using the `path_out` parameter.

### Using `cross_plot` 

This function shows the distribution of both, numerical and categorical variables, when predicting a binary (or two-class), target variable.

The two plots shows the same input data, but in different ways.

```{r}
vars_to_process_2=c("thal", "max_heart_rate")
cross_plot(data=heart_disease, str_input=vars_to_process_2, str_target="has_heart_disease")
```


If the input variable is numerical and it has more than 20 unique values, then it will automatically creates bins or buckets based on equal frequency criteria to better show the relationship. `max_heart_rate` is an example of this behavior. 

It can be disabled using `auto_binning=FALSE` parameter.

```{r, eval=FALSE}
# It will run for all variables (str_input is missing) and it will export all the plots.
cross_plot(data=heart_disease, str_target="has_heart_disease", path_out = "my_folder")
```

### Using `plotar`

```{r}
plotar(data=heart_disease, str_input = 'age', str_target="has_heart_disease", plot_type="histdens")
plotar(heart_disease, str_input = 'age', str_target = 'has_heart_disease', plot_type = "boxplot")
```

